Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-11 write-ups/PLDI/paper.tex
Tweaks.

We provide background on the industry-standard OpenCL programming
model on which we base cooperative kernels (\mysec\ref{sec:opencl}),
and illustrate OpenCL using two irregular data-parallel algorithms---a
work stealing queue and breadth-first graph search---that also
pinpoint the problems associated with unfair scheduling of workgroups
and motivate our cooperative threads approach (\mysec\ref{sec:openclexamples}.

@@

\myfiglong\ref{fig:workstealing} shows a simplified version of an OpenCL
kernel that uses a work stealing queue. All threads start their execution at the
function marked with the $\mathsf{kernel}$ keyword. They receive a
pointer to the task pools, which reside in global memory, and which
were initialized by the host program beforehand to contain the initial
tasks. Each thread can identify the workgroup it belongs to via the
function $\getgroupid()$.  Here the group id is stored into the
variable $\mathsf{pool\_id}$ (which is, by default, in private memory)
and used to access the relevant task pool. The
$\mathsf{pop\_or\_steal}$ function is called by all threads of a
workgroup.  Inside $\mathsf{pop\_or\_steal}$ only the thread with a local id equal to
$0$, also called the \emph{master thread} of the workgroup,
\ADComment{Hugues, I changed; please check:} will try to pop a task
from the workgroup's pool, or steal a task from a different pool. The
master thread uses local memory inside $\mathsf{pop\_or\_steal}$ to
ensure that the returned value is valid for all threads of its
workgroup. If a task was obtained, then all threads of the workgroup
participate in the processing of the task via the
$\mathsf{process\_task}$ function. \HEComment{Manage to say that a
  task is considered big enough to benefit from parallel computation}
\ADComment{I think that would be too much detail given the space we
  have, so I would not delve into this. Remove both comments if you
  agree.} Processing this task may lead to the creation of further
tasks, which the master thread will push to the workgroup's pool.

@@

To achieve this balance, both an implementation of the cooperative
kernels model, and the programmer of a cooperative kernel, must strive
to meet a number of non-functional requirements, which we outline
here.

\ADComment{Sree mentioned TimeGraph paper.}

@@

Recall that the purpose of $\offerkill$ is to provide the scheduler
with an opportunity to destroy a workgroup in order to schedule
higher-priority tasks.  Because these higher-priority tasks, such as
graphics rendering, may have soft real-time constraints, the scheduler
relies on the cooperative kernel to execute $\offerkill$ sufficiently
frequently that the scheduler can meet such real-time constraints.

\ADComment{Forward reference to experiments with display freezing.}

@@

\ADComment{Generosity: initial number of workgroups, and responding to offer fork.}

>>>>

We provide background on the industry-standard OpenCL programming
model on which we base cooperative kernels (\mysec\ref{sec:opencl}),
and illustrate OpenCL using two irregular data-parallel algorithms---a
work stealing queue and breadth-first graph search---that also
pinpoint the problems associated with unfair scheduling of workgroups
and motivate our cooperative threads approach (\mysec\ref{sec:openclexamples}).

@@

\myfiglong\ref{fig:workstealing} shows a simplified version of an OpenCL
kernel that uses a work stealing queue. All threads start their execution at the
function marked with the $\mathsf{kernel}$ keyword. They receive a
pointer to the task pools, which reside in global memory, and which
were initialized by the host program beforehand to contain the initial
tasks. Each thread can identify the workgroup it belongs to via the
function $\getgroupid()$.  Here the group id is stored into the
variable $\mathsf{pool\_id}$ (which is, by default, in private memory)
and used to access the relevant task pool. The
$\mathsf{pop\_or\_steal}$ function is called by all threads of a
workgroup.  Inside $\mathsf{pop\_or\_steal}$ only the thread with a local id equal to
$0$, also called the \emph{master thread} of the workgroup,
will try to pop a task
from the workgroup's pool, or steal a task from a different pool. The
master thread uses local memory inside $\mathsf{pop\_or\_steal}$ to
ensure that the returned value is valid for all threads of its
workgroup. If a task was obtained, then all threads of the workgroup
participate in the processing of the task via the
$\mathsf{process\_task}$ function.  Processing this task may lead to the creation of further
tasks, which the master thread will push to the workgroup's pool.

@@

To achieve this balance, both an implementation of the cooperative
kernels model, and the programmer of a cooperative kernel, must strive
to meet the following non-functional requirements.

@@

The purpose of $\offerkill$ is to provide the scheduler with an
opportunity to destroy a workgroup in order to schedule
higher-priority tasks.  Because these higher-priority tasks, such as
graphics rendering, may have soft real-time constraints, the scheduler
relies on the cooperative kernel to execute $\offerkill$ sufficiently
frequently that the scheduler can meet such real-time constraints.

Using work stealing as an example: in the kernel of \myfig\ref{...}, a
workgroup offers itself to the scheduler after processing each task.
If tasks turn out to be very short then it might turn out to be
reasonable to invoke $\offerkill$ less frequently, after every $k$
tasks for some appropriate $k$.  On the other hand, if individual
tasks may be very time-consuming then it may be necessary to rewrite
the algorithm so work on a larger number of shorter tasks, so as to
achieve a higher rate of calls to $\offerfork$.

Getting this non-functional requirement right depends on the
constraints of the tasks to be co-scheduled with a cooperative kernel,
on properties of the cooperative kernel itself, and on the efficiency
of the GPU that is used for execution.  In \mysec\ref{...} we conduct
experiments to understand the response rate that would be required to
co-schedule graphics rendering with a cooperative kernel in order to
achieve a smooth frame rate, and we evalaute responsiveness in
practice in \mysec\ref{...}.

\paragraph{Sufficient workgroups for the cooperative kernel}

Recall that the cooperative kernel would ideally utilise the $N$
workgroups requested upon kernel launch, in order to conduct its
processing efficiently.  The scheduler should thus aim to provide with
cooperative kernel with as many workgroups, up to this number, as
other constraints allow.  As a result, the scheduler should generally
accept an $\offerkill$ only if a comput unit is required for another
task, and should respond positively to $\offerfork$ calls by
scheduling new workgroups if compute units are available.  Our
implementation of cooperative scheduling (\mysec\ref{ }) is as
generous as possible to the cooperative kernel in this regard.
However, these are not hard-and-fast rules: a scheduler might employ
heuristics based on historical workloads to avoid thrashing.  For
example, if graphics calls are known to arrive at a certain rate then
the scheduler might decide that it is worth accepting an $\offerkill$
from the cooperative kernel on the assumption that a graphics task
will be requested momentarily.  Similarly, the scheduler might provide
some, but not all, compute units to the cooperative kernel in response
to an $\offerfork$ call if it has reason to believe that the remaining
compute units are likely to be useful for other high priority tasks
that are likely to arrive imminently.

\paragraph{Killing order}

We opted for a semantics whereby only the active workgroup with the
highest id can be killed.  This has an appealing property: it means
that the ids of active workgroups are contiguous, which is important
for processing of contiguous data.  The cooperative graph traversal
algorithm of \myfig\ref{ } illustrates this: the algorithm is prepared
for $\getglobalsize()$ to change after each resizing barrier call, but
depends on the fact that $\getglobalid()$ returns a contiguous range
of thread ids.

@@

\paragraph{Keeping one workgroup alive}
